---
title: "Analysis of Storm Data: Measuring the impact of weather events"
author: "Julian Cuero"
date: "12/11/2020"
output: html_document
---

## Synopsis
The following analysis is about the consequences of severe weather events in the United States during the period that goes from 1950 to 2011. The data comes from the National Weather Service (NWS) Storm Data, which consists of 37 variables and 902297 observations. Accordingly, the NWS does not guarantee the accuracy or validity of the information that will be presented here. However, the following analysis might be useful to prioritize resources for different types of events.

For additional details you can check the [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

## Setup
```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
```

## Research Questions
As could have been suggested from above, the research questions are outlined in the following way:

1. Across the United States, which types of events
are most harmful with respect to population health?
2. Across the United States, which types of events
have the greatest economic consequences?

## Loading data
```{r, cache=TRUE}
raw_data <- read.csv("StormData.csv", header=TRUE)
```

## Data Processing
Because of the bulk of data and guided by the the research questions, only relevant variables will be
selected. In this order of ideas, these are the chosen variables from the original dataset:

* `EVTYPE`: Represents the event types, which represent the main variable of interest.
* `INJURIES`: Number of injuries of an event in a given date.
* `FATALITIES`: Number of fatalities of an event in a given date.
* `CROPDMG`: Crop damage.
* `CROPDMGEXP`: Crop damage exponent.
* `PROPDMG`: Property damage.
* `PROPDMGEXP`: Property damage exponent.

```{r}
new_data <- raw_data[,c("EVTYPE","INJURIES", "FATALITIES", "CROPDMG", "CROPDMGEXP", "PROPDMG", "PROPDMGEXP")]
```

Then, the columns are renamed, so they are a bit easier to access.

```{r}
names(new_data) <- tolower(names(new_data))
```

After that, it's necessary to do some cleanup for three variables (`EVTYPE`, `CROPDMGEXP` and `PROPDMG`).

To begin with, according to the NWS docs, in *Table 1 of Section 2.1.1*,  there are **48 event types**:

1. Astronomical Low Tide
2. Avalanche
3. Blizzard
4. Coastal Flood
5. Cold/Wind Chill
6. Debris Flow
7. Dense Fog
8. Dense Smoke
9. Drought
10. Dust Devil
11. Dust Storm
12. Excessive Heat
13. Extreme Cold/Wind Chill
14. Flash Flood
15. Flood
16. Freezing Fog
17. Frost/Freeze
18. Funnel Cloud
19. Hail
20. Heat
21. Heavy Rain
22. Heavy Snow
23. High Surf
24. High Wind
25. Hurricane/Typhoon
26. Ice Storm
27. Lakeshore Flood
28. Lake-Effect Snow
29. Lightning
30. Marine Hail
31. Marine High Wind
32. Marine Strong Wind
33. Marine Thunderstorm Wind
34. Rip Current
35. Seiche
36. Sleet
37. Storm Tide
38. Strong Wind
39. Thunderstorm Wind
40. Tornado
41. Tropical Depression
42. Tropical Storm
43. Tsunami
44. Volcanic Ash
45. Waterspout
46. Wildfire
47. Winter Storm
48. Winter Weather

However, when looking at the data, there are **985** different event types because they were entered differently.

```{r, cache=TRUE}
length(unique(raw_data$EVTYPE))
```

Therefore, I performed some string processing. The way I classified a specific entry to an event category was the one that most accurately describes the meteorological event. If a specific category does not appear in the code is because I did not find any anomalies in the entries related to it.

```{r, cache=TRUE}
data_clean <- new_data %>% mutate(evtype=case_when(
    grepl("astro", evtype, ignore.case = TRUE) ~ toupper("Astronomical Low Tide"),
    grepl("^aval", evtype, ignore.case = TRUE) ~ toupper("Avalanche"),
    grepl("^blizzard", evtype, ignore.case = TRUE) ~ toupper("Blizzard"),
    grepl("^[ ]*coastal", evtype, ignore.case = TRUE) ~ toupper("Coastal Flood"),
    grepl("Record cold|^cold", evtype, ignore.case = TRUE) ~ toupper("Cold/Wind Chill"),
    grepl("^freezing Fog", evtype, ignore.case = TRUE) ~ toupper("Freezing Fog"),
    grepl("fog", evtype, ignore.case = TRUE) ~ toupper("Dense Fog"),
    grepl("smoke", evtype, ignore.case = TRUE) ~ toupper("Dense Smoke"),
    grepl("^drought|EXCESSIVELY DRY", evtype, ignore.case = TRUE) ~ toupper("Drought"),
    grepl("devil", evtype, ignore.case = TRUE) ~ toupper("Dust Devil"),
    grepl("^dust storm", evtype, ignore.case = TRUE) ~ toupper("Dust Storm"),
    grepl("^EXCESSIVE HEAT|RECORD/EXCESSIVE HEAT", evtype, ignore.case = TRUE) ~ toupper("Excessive Heat"),
    grepl("^Extreme Cold|^prolong cold", evtype, ignore.case = TRUE) ~ toupper("Extreme Cold/Wind Chill"),
    grepl("LOCAL FLASH FLOOD|FLOOD FLASH|Flash Flood", evtype, ignore.case = TRUE) ~ toupper("Flash Flood"),
    grepl("^Flood|Breakup flooding|River flood", evtype, ignore.case = TRUE) ~ toupper("Flood"),
    grepl("^Frost|^Freeze", evtype, ignore.case = TRUE) ~ toupper("Frost/Freeze"),
    grepl("^Funnel|funnels$|funnel$", evtype, ignore.case = TRUE) ~ toupper("Funnel Cloud"),
    grepl("^hail|small hail|NON SEVERE HAIL|deep hail", evtype, ignore.case = TRUE) ~ toupper("Hail"),
    grepl("^heat|^record heat", evtype, ignore.case = TRUE) ~ toupper("Heat"),
    grepl("^Heavy rain|^FREEZING RAIN|rainfall|^rain|^HVY RAIN|TORRENTIAL RAIN", evtype, ignore.case = TRUE) ~ toupper("Heavy Rain"),
    grepl("^snow|^Heavy snow|^BLOWING SNOW|^Record May Snow|^Light snow|^MODERATE SNOW|^FALLING SNOW|^Mountain Snows|^Thundersnow|late snow", evtype, ignore.case = TRUE) ~ toupper("Heavy Snow"),
    grepl("^heavy surf|^high surf|HAZARDOUS SURF|^[ ]*HIGH SURF|^ROUGH SURF", evtype, ignore.case = TRUE) ~ toupper("High Surf"),
    grepl("^high wind|Gusty wind|GUSTY LAKE WIND", evtype, ignore.case = TRUE) ~ toupper("High Wind"),
    grepl("typhoon|^hurricane", evtype, ignore.case = TRUE) ~ toupper("Hurricane/Typhoon"),
    grepl("^Ice|GLAZE ICE|^Glaze|PATCHY ICE|^Black Ice", evtype, ignore.case = TRUE) ~ toupper("Ice Storm"),
    grepl("^LAKESHORE", evtype, ignore.case = TRUE) ~ toupper("Lakeshore Flood"),
    grepl("^LAKE FLOOD|^Lake-Effect|^HEAVY LAKE SNOW", evtype, ignore.case = TRUE) ~ toupper("Lake-Effect Snow"),
    grepl("^[ ]*Lightning", evtype, ignore.case = TRUE) ~ toupper("Lightning"),
    grepl("Marine Hail", evtype, ignore.case = TRUE) ~ toupper("Marine Hail"),
    grepl("^rip current", evtype, ignore.case = TRUE) ~ toupper("Rip Current"),
    grepl("^sleet", evtype, ignore.case = TRUE) ~ toupper("Sleet"),
    grepl("^STORM SURGE|^HIGH TIDES|^BLOW-OUT TIDE", evtype, ignore.case = TRUE) ~ toupper("Storm Tide"),
    grepl("^Strong wind", evtype, ignore.case = TRUE) ~ toupper("Strong Wind"),
    grepl("GUSTY THUNDERSTORM WIND|^Thunderstorm Wind|^THUNDERSTORM|^SEVERE THUNDERSTORM|^[ ]*TSTM", evtype, ignore.case = TRUE) ~ toupper("Thunderstorm Wind"),
    grepl("^tornado", evtype, ignore.case = TRUE) ~ toupper("Tornado"),
    grepl("^Tropical Storm", evtype, ignore.case = TRUE) ~ toupper("Tropical Storm"),
    grepl("volcanic", evtype, ignore.case = TRUE) ~ toupper("Volcanic Ash"),
    grepl("^Waterspout|^[ ]*Waterspout", evtype, ignore.case = TRUE) ~ toupper("Waterspout"),
    grepl("Wildfire", evtype, ignore.case = TRUE) ~ toupper("Wildfire"),
    grepl("^Winter storm|^Record Winter Snow", evtype, ignore.case = TRUE) ~ toupper("Winter Storm"),
    grepl("^Winter Mix|^Wintery Mix|^WINTER WEATHER", evtype, ignore.case = TRUE) ~ toupper("Winter Weather"),
    TRUE ~ evtype
))
```

Although I did not reduced the 985 different event types to 48, I reduced it by more than half (438 specifically). It must be highlighted that I didn't do so because the purpose of the project is about reproducible research and data analysis, not just data processing. Besides that, it was difficult to cover all possible cases.

```{r}
length(unique(data_clean$evtype))
```

Second, I also found necessary to clean the `PROPDMGEXP` and `CROPDMGEXP` variables. Because
these are the exponents/multipliers for `PROPDMG` and `CROPDMG`, which will serve as a measure of the economic impact of weather events.

These are the possible values of `CROPDMGEXP` and `PROPDMGEXP`:
**H,h,K,k,M,m,B,b,+,-,?,0,1,2,3,4,5,6,7,8,** and **blank-character**

```{r}
unique(raw_data$PROPDMGEXP)
```

According to the National Oceanic and Atmospheric Administration's (NOAA) Storm Database, I decided to classify those values as follows:

- H,h = hundreds = 100
- K,k = kilos = thousands = 1,000
- M,m = millions = 1,000,000
- B,b = billions = 1,000,000,000
- numeric (0-8) = 10
- (+) = 1
- (-) = 0
- (?) = 0
- empty character ("") = 0

Thus, this is what the `replace_num` function does, it replaces the character with its respective numeric value.

```{r}
replace_num <- function(vec){
    nums <- as.character(0:8)
    for(i in 1:length(vec)){
        if(tolower(vec[i]) == "k") vec[i] <- 1000
        else if(tolower(vec[i]) == "m") vec[i] <- 1000000
        else if(vec[i] %in% nums) vec[i] <- 10
        else if(tolower(vec[i]) == "h") vec[i] <- 100
        else if(tolower(vec[i]) == "b") vec[i] <- 1000000000
        else if(vec[i] == "+") vec[i] <- 1
        else {
            vec[i] <- 0
        }
    }
    as.integer(vec)
}
```

And now, to calculate economic impact I created
two more variables: `property_damage` and `crop_damage`:

- `property_damage` = `propdmgexp` * `propdmg`
- `crop_damage` = `cropdmgexp` * `cropdmg`

```{r}
data_clean <- data_clean %>% mutate(propdmgexp=replace_num(propdmgexp),
                cropdmgexp=replace_num(cropdmgexp),
                crop_damage=cropdmg * cropdmgexp,
                property_damage=propdmg * propdmgexp)
```

## Data analysis
#### Research question 1
In order to quantify health consequences,
total **injuries** and total **fatalities** will serve for this purpose. Since we're only interested in the ones that affected health the most and there's quite a lot of data it's helpful to filter events that contain at least 1 injury or 1 fatality.

```{r}
harmful <- subset(data_clean, injuries > 0 | fatalities > 0)
```

As a result, **total fatalities and injuries** are calculated **by event type** and we just get the **top 6**.

```{r, message=FALSE}
most_harmful <- harmful %>% group_by(evtype) %>%
    summarise(total_injuries=sum(injuries),             total_fatalities=sum(fatalities)) %>%
    arrange(desc(total_fatalities)) %>%
    top_n(6)

most_harmful
```

##### Fatalities
In this first plot we start seeing that **tornadoes** are the event that unfortunately kill people the most, followed by excessive heat and flash flood. Suggesting that these may be the events that harm American health the most.

```{r}
most_harmful %>% ggplot(aes(x=reorder(evtype, -total_fatalities),y=total_fatalities,fill=evtype)) +
    geom_bar(stat="identity") + labs(x="Event type", y="Total fatalities") +
    scale_fill_brewer(palette = "RdGy") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
```

##### Injuries
Similarly, **tornadoes** occupy the first place in the total number of injuries, and we can also find excessive heat and thunderstorm wind, which seems to confirm the previous findings.

```{r}
most_harmful %>% ggplot(aes(x=reorder(evtype, -total_injuries),y=total_injuries,fill=evtype)) +
    geom_bar(stat="identity") + labs(x="Event type", y="Total injuries") +
    scale_fill_brewer(palette = "RdGy") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
```

#### Research question 2
In regard to the second question, I calculated `total_crop_damage` and `total_prop_damage`, where the first is the sum of `crop_damage` measured in billions and the second is also the sum of `property_damage` in billions.

```{r, message=FALSE}
worst_econ <- data_clean %>% group_by(evtype) %>%
    summarise(total_crop_damage=sum(crop_damage)/10^9, total_prop_damage=sum(property_damage)/10^9) %>%
    arrange(desc(total_prop_damage)) %>%
    top_n(5)
worst_econ
```

In this final plot, we now find that **floods** top the list, then there are **hurricanes or typhoons** followed by **tornadoes** which caused property damage of around 150 billions, 85.4 billions and 58.5 billions respectively.

```{r}
worst_econ_plot <- with(worst_econ,
                data.frame(evtype=c(evtype,evtype),               total=c(total_prop_damage,total_crop_damage),
type=c(rep("Property damage",nrow(worst_econ)),
       rep("Crop damage",nrow(worst_econ)))))
```

```{r}
worst_econ_plot %>% ggplot(aes(x=reorder(evtype, -total),y=total,fill=type)) +
    geom_bar(stat="identity",position = "dodge") +
    expand_limits(x = c(0, NA), y = c(0, NA)) +
    scale_y_continuous(labels = unit_format(unit = "B")) +
    labs(x="Event type",y="Total losses (in billions)") +
    scale_fill_manual(values=c("#af2d2d","#f05454")) + 
    coord_flip()
```

## Results
Finally, I obtained two conclusions based on the
research questions:

1. As we saw above, if measured by fatalities, the event that damaged American health the most in this period were the **tornadoes**, followed by the **heat** in general (regular and excessive heat), and **flash flood**. If we also see the number of injuries, we can confirm that tornadoes are the most significant threat to health as well as excessive heat, but thunderstorm wind should also be regarded.

2. Measured by property and crop damage, the types of events that have the most negative economic impact across the USA are **floods**, **hurricanes or typhoons** and, once again, **tornadoes**.